Are populations like a circuit? Comparing isolation by resistance to a new coalescent-based method.

### 2019-04-19 review-responses.tex

\begin{quote}
    This article focuses on the inference of gene flows between populations from
    the analysis of georeferenced genetic data. In particular, a new
    coalescent-based technique, which applies to the analysis of pairs of
    sampled individuals, is compared to an approach based on "resistance
    distances" which aim at approximating coalescent distances. This work is
    interesting because resistance distances are at the core of several popular
    techniques in landscape genetics. It is thus important to gain a good
    understanding of how good this phenomenological approximation is compared
    to more mechanistic techniques such as the structured coalescent. The
    article clarifies the connections between the two approaches in a
    convincing manner. It thereby fully deserves to be published in Molecular
    Ecology Resources. There are however a few points (see below) that I think
    should be improved should the authors see fit.
\end{quote}

% Section "Hitting times of Markov chains"

\begin{point}{}
    $Pr(X_{t+e}=y | X_t=y)$ should read $Pr(X_{t+e}=y | X_t=x)$.
\end{point}

\reply{
    Whoops, thanks! \revref
}

\begin{point}{}
    Regarding Equation 2, I would recommend giving more detail explaining where
    this comes from. Here is a suggestion:
    \begin{align*}
        H_{xz} &= dt + \sum_y Pr(X_{dt} = y | X_0 = x) H_{yz} \\
               &= dt + \sum_{y != x} Q_{xy} dt H_{yz} + (Q_{xx} dt + 1) H_{xz}
    \end{align*}
\end{point}

\reply{
    We wondered whether this would overload readers,
    but have now included an explanation along these lines.
    \revref
}

\begin{point}{}
    On a related note, it might be helpful to give more explanations about Equation 6. In
    fact, I am not quite sure I agree with it. Indeed, for $x \neq z$, we have
    \begin{align*}
        C_{xz} &= dt + {} \\
                & \qquad
             1/2 \sum_y Pr(X_{dt} = y | X_0 = x) C_{yz} + {} \\
                & \qquad
             1/2 \sum_y Pr(X_{dt} = y | X_0 = z) C_{xy}, \\
           &= dt + {} \\
                & \qquad
             1/2 \sum_{y \neq x} Q_{xy} dt C_{yz} + (Q_{xx} dt + 1) C_{xz} + \\
                & \qquad
             1/2 \sum_{y \neq x} Q_{zy} dt C_{xy} + (Q_{xx} dt + 1) C_{xz},
    \end{align*}
    where the 1/2 terms correspond to the probability of "selecting" the lineage found in x
    (resp. z) at time 0 to move to y after dt. The last equation above then gives
    \begin{align*}
        -2 = \sum_y Q_{xy} C_{yz} + Q_{zy} C_{xy}
    \end{align*}
    which differs from the one given in the ms. I have probably missed something in the derivation
    used by the authors. Giving a few more explanations here might help avoiding any confusion.
\end{point}

\reply{
    Besides substituting $G$ for $Q$,
    our equations only differ by the factor of 1/2.
    The model we work under is that each lineage moves as an independent copy
    of the Markov chain $X$, so that the total jump rate
    when the two chains are at $x$ and $z$ is $-G_{xx}-G_{zz}$.
    We think the factors of 1/2 are in error --
    a more clear way to see this would be to say that $X$ and $Z$ are the locations of the two lineages,
    that move independently until they coalesce,
    and so
    \begin{align*}
        C_{xz} &= dt + {} \\
                & \qquad
             \sum_u \P(X_{dt} = u | X_0 = x) C_{uz} + {} \\
                & \qquad
             \sum_v \P(Z_{dt} = v | Z_0 = z) C_{xv}, \\
           &= dt + {} \\
                & \qquad
             \sum_{u \neq x} G_{xu} dt C_{uz} + (G_{xx} dt + 1) C_{xz} + \\
                & \qquad
             \sum_{v \neq x} G_{zv} dt C_{xv} + (G_{xx} dt + 1) C_{xz} .
    \end{align*}
}

@@

% Section "Bayesian inference of movement rates"

\begin{point}{}
    I was also a bit puzzled by the likelihood function that is at the core of the
    Bayesian inference proposed here (see Equation 8). The spatial component of
    the data seems completely hidden. It is in fact conveyed by the coalescent
    times $\mathcal{C}_{ij}(G,\gamma)$ but a more detailed description of the likelihood
    function where both genetic and spatial information appear explicitly would
    facilitate the understanding of the proposed inference technique.
\end{point}

@@

\reply{
    We've added a clarifying sentence. \revref
}

\begin{point}{}
    Moreover, my understanding is that the joint probability of the pairwise genetic distances
    ($D$) and spatial locations ($L$) given $G$ and $\gamma$, i.e.,
    $p(D,L|G,\gamma)$, is expressed as the product 
    $p(D|L,G,\gamma) \times p(L|G,\gamma)$.  Thanks to Equation (7) the first density is in fact
    $p(D|C)$, where $C$ are the coalescent times. The authors use a normal distribution for $D|C$ but
    nothing is said  about the distribution of $L|G,\gamma$. It is paramount to explicitly define
    every probabilistic distribution in the likelihood of the model used for the inference so as
    to clarify what are the underlying simplifying hypotheses the proposed approach relies on.
\end{point}

\reply{
    In our model, spatial locations are (discrete and) assumed to be known without error,
    and we do not want to model the sampling process,
    so they do not enter into the likelihood.
    Also recall that both $D$ and $C$ are \emph{complete} matrices of genetic distances 
    and pairwise coalescence times -- i.e.,
    between all pairs of populations --
    so the locations do not enter at this stage (although they are certainly used in estimating $D$).
    We have added a note reminding the reader of this. \revref
}

@@

\begin{point}{}
    On a related note, my understanding is that the uncertainty affecting
    coalescent times conditional on the movement rates is not taken into
    account in the proposed approach (i.e., knowing $G$, one gets $C$ through
    Equation 7 in a deterministic fashion). I would recommend that the authors
    discuss this point.
\end{point}

\reply{
    This is true; we've added a sentence emphasizing this. \revref
}

@@

% Section "Identifying a barrier to gene flow"

\begin{point}{}
    The sentence ``with noise standard deviation equal to 1/1000th of the mean
    coalescence time (left) and 1/100th of the mean coalescence time (right)''
    should read ``with noise standard deviation equal to 1/1000th of the mean
    coalescence time (top) and 1/100th of the mean coalescence time (bottom)''
\end{point}

\reply{
    Whoops! Fixed. \revref
}

@@

% Section "Discussion"

\begin{point}{}
    The authors call for new models to be developed that would properly accommodate
    for individuals distributed in space along a continuum as opposed to
    discrete demes, as is the case with the structured coalescent. The spatial
    Lambda-Fleming Viot model is one of these. It has been studied extensively
    from a mathematical standpoint (see the body of work by Etheridge, Barton
    and Veber amongst others) and has been used for the inference using a
    Bayesian approach (see Guindon, Guo, Welch, Theoretical Population Biology,
    2016; but also Koskela's PhD thesis
    \url{http://wrap.warwick.ac.uk/88065/1/WRAP_Theses_Koskela_2016.pdf}).  The
    spatial Lambda-Fleming-Viot model is mechanistic and enjoys many good
    statistical and biological properties (e.g., it does not suffer from the
    ``clumping'' issue that affects the Wright-Malecot model).
\end{point}

\reply{
    Certainly! We agree that the SLFV model is a promising way forward
    to be able to use coalescent simulations in continuous space.
    There's a lot of work left to explore how and when it's a good approximation
    to reality also,
    but we've added a bit about it to the ``Other methods'' section. \revref
}

@@

\begin{point}{}
    Assuming that space is discrete in arguably the biggest weakness of the structured
    coalescent. It is quite often not biologically realistic and breaks down
    mathematically when considering larger numbers of smaller demes. The
    spatial Lambda-Fleming-Viot avoids this important shortcoming and thus
    probably deserves more attention.
\end{point}

\reply{
    We agree, and very much hope to move in that direction.
    We've added more discussion of this. \revref
}

### 2019-04-19 isolation_by_coalescence.tex
done going through reviews

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the ``jump rate'' from $x$ to $y$.
The jump rate $G_{xy}$ is \revpoint{3}{8}
the probability the chain jumps to position $y$ when at position $x$ per unit of time.
(More precisely, if the Markov chain is at location $X_t$ at time $t$, 
then for $x \neq y$, $\PP{X_{t+\epsilon} = y \given X_t = x} = \epsilon G_{xy} + O(\epsilon^2)$.)
For mathematical convenience, the diagonal of this matrix is completed so that rows sum to zero:
$G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain --
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
By conditioning on $X_\epsilon$,
one can use the Markov property to show that
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z,
\end{align}
i.e., if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$ \citep{kemeny1983finite}.

@@

Mean coalescence times estimated from real data are subject to a number of sources of noise
(which we discuss in more detail later).
Here, we use
exact solutions of the equations above
to develop a Bayesian inference method that accounts for noise in the data
and estimates uncertainty in the resulting estimates.
To do this, we model genetic distances as equal to coalescence times plus noise:
$D_{ij} = C_{ij} + \epsilon_{ij}$,
where $\epsilon_{ij}$ are independent and Normally distributed
with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
except that $\epsilon_{ij} = \epsilon_{ji}$.
(We assume that coalescence times are measured to good accuracy 
-- i.e., $C_{ij}$ is large compared to $\sigma_{\epsilon}$ -- \revpoint{3}{3}
so that this model does not predict negative times.)
Given $G$ and $\gamma$ we can solve equation \eqref{eqn:C_matrix} 
to find the corresponding coalescence times
-- which we call $\mathcal{C}(G, \gamma)$ --
so our model is that $D$ has a Gaussian distribution 
with mean $\mathcal{C}(G, \gamma)$.
The maximum likelihood estimate of $G$ could be found directly;
however, this in practice will quite likely contain negative movement rates,
and does not allow us to impose constraints
(such as only allowing a subset of movement rates to be nonzero).

Therefore, we use Bayesian inference,
placing independent exponential priors on the parameters --
the movement rates, $G_{ij}$, that are not constrained to be zero
and the coalescence rates, $\gamma$.
Concretely, suppose that the spatial arrangement of populations is given by a graph 
(i.e., a discrete collection of locations represented by ``nodes'',
between which movement is possible only between those connected by ``edges''). \revpoint{3}{24}
Only movement rates in $G$ that correspond to edges in the graph are allowed to be nonzero.
Suppose there are $m_G$ edges in the graph,
that each of the $n$ populations has its own coalescence rate,
and write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph.
Then, the resulting log-posterior density is
\begin{align} \label{eq:post}
    \begin{split}
\mathcal{L}(D, G, \gamma) 
    &=
    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
	- \sum_{i \sim j} \lambda_G G_{ij}  \\
    &\qquad
    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
        \right)^2 
    \end{split}
\end{align}
Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
respectively.

@@

We then test the method's ability to locate barriers to migration
on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
Each nonzero movement rate is determined randomly as before,
and coalescence rates are set to 1 everywhere.
Figure \ref{fig:5x3b_post_coalvcom} shows posterior distributions of the movement rates
for both coalescence time and resistance distance inference
on this graph, with noise standard deviation equal to
1/1000${}^\text{th}$ of the mean coalescence time (left)
and 1/100${}^\text{th}$ of the mean coalescence time (right).
Although increased noise in the data increased uncertainty in the estimated values,
coalescence time inference correctly inferred not only the location of the barrier
but also the migration rates along all other edges within a reasonable margin of error.
Resistance distance inference also identified the barrier,
but was much less accurate for the other values.
We also explored making the problem harder,
by first (a) dropping data corresponding to populations 2, 5, 7, and 15;
and then (b) allowing a separate coalescence rate for each location,
in both cases using a noise level of 1/100.
The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
and show substantially increased error,
but only in movement rates to the removed locations.
As before, uncertainty was increased when multiple coalescence rates were allowed,
but not as strongly as before, likely because the problem is less close to being underconstrained,
having 105 equations and 59 unknowns.

@@

Most promisingly, the recent method, MAPS \citep{alasadi2018estimating}
is based on EEMS, but instead of resistance distances,
uses pairwise sharing of long haplotype segments,
which have been shown to carry substantial information about recent demography
\citep{ralph2013geography,palamara2013inference,browning2015accurate,ringbauer2017inferring}.
Furthermore, the likelihood model underlying MAPS is based on coalescent theory,
and thus should not be subject to the drawbacks of resistance distance.
\citet{ringbauer2018estimating} also uses shared haplotype lengths
to identify a barrier in a coalescence-based framework.
However, the underlying migration model is symmetric,
and haplotype length-based methods require fairly dense genotyping and a genetic map,
something not available for many species.

@@

>>>>

Now, we explain how to compute the relevant quantities --
mean commute and coalescence times -- from a Markov chain model of lineage movement.
The Markov chain is defined by its \emph{generator matrix}, denoted $G$,
for which $G_{xy}$ gives the ``jump rate'' from $x$ to $y$.
The jump rate $G_{xy}$ is \revpoint{3}{8}
the probability the chain jumps to position $y$ when at position $x$ per unit of time.
(More precisely, if the Markov chain is at location $X_t$ at time $t$,  \revpoint{5}{1}
then for $x \neq y$, $\PP{X_{t+\epsilon} = y \given X_t = x} = \epsilon G_{xy} + O(\epsilon^2)$.)
For mathematical convenience, the diagonal of this matrix is completed so that rows sum to zero:
$G_{xx} = - \sum_{y \neq x} G_{xy}$.
We will need to find the \emph{hitting times} of the chain --
i.e., for each pair of states $x$ and $y$, 
the mean time until the chain first hits $y$ after being started at $x$.
We denote this quantity $H_{xy}$.
Throughout, we assume that all hitting times are finite,
which implies the chain is connected and irreducible.
The Markov property lets us write a system of linear equations that can be solved for $H$, as follows
(see, e.g., \citet{kemeny1983finite}):
Suppose that the chain begins at $x \neq z$,
and is at location $y \neq z$ after $dt$ units of time, without having visited $z$.
Then the total time to hit $z$ is equal to the elapsed time $dt$
plus the remaining time to hit $z$ from $y$.
Formally, this says that \revpoint{5}{2}
$H_{xz} \approx dt + \sum_y \P\{X_{dt} = y \given X_0 = x\} H_{yz}$.
Replacing $\P\{X_{dt} = y \given X_0 = x\}$ with $dt G_{xy}$ and rearranging
gives the system of equations:
\begin{align} \label{eqn:hitting_sum}
    \sum_y G_{xy} H_{yz} = -1 \qquad \text{for} \; x \neq z.
\end{align}
In other words, if $G_{-z}$ is the matrix with the $z^\text{th}$ row and column removed,
and $\bone$ is the vector composed of all ones,
then the $z^\text{th}$ column of $H$, except for $H_{zz}$,
can be computed as $- (G_{-z})^{-1} \bone$.
This, along with $H_{zz} = 0$, allows computation of $H$
from the movement rates $G$ \citep{kemeny1983finite}.

@@

Mean coalescence times estimated from real data are subject to a number of sources of noise
(which we discuss in more detail later).
Here, we use
exact solutions of the equations above
to develop a Bayesian inference method that accounts for noise in the data
and estimates uncertainty in the resulting estimates.
To do this, we model genetic distances as equal to coalescence times plus noise:
$D_{ij} = C_{ij} + \epsilon_{ij}$,
where $\epsilon_{ij}$ are independent and Gaussian distributed
with mean $\mathbf{0}$ and variance $\sigma_\epsilon^2$,
except that $\epsilon_{ij} = \epsilon_{ji}$.
(We assume that coalescence times are measured to good accuracy 
-- i.e., $C_{ij}$ is large compared to $\sigma_{\epsilon}$ -- \revpoint{3}{3}
so that this model does not predict negative times.)
Given $G$ and $\gamma$ we can solve equation \eqref{eqn:C_matrix} 
to find the corresponding coalescence times
-- which we call $\mathcal{C}(G, \gamma)$ --
so our model is that $D$ has a Gaussian distribution 
with mean $\mathcal{C}(G, \gamma)$.
(This Gaussian stands in for the sampling noise of coalescence times about their mean.) \revpoint{5}{6}
The maximum likelihood estimate of $G$ could be found directly;
however, this in practice will quite likely contain negative movement rates,
and does not allow us to impose constraints
(such as only allowing a subset of movement rates to be nonzero).

Therefore, we use Bayesian inference,
placing independent exponential priors on the parameters --
the movement rates, $G_{ij}$, that are not constrained to be zero
and the coalescence rates, $\gamma$.
Concretely, suppose that the spatial arrangement of populations is given by a graph 
(i.e., a discrete collection of locations represented by ``nodes'',
between which movement is possible only between those connected by ``edges''). \revpoint{3}{24}
Only movement rates in $G$ that correspond to edges in the graph are allowed to be nonzero.
Suppose there are $m_G$ edges in the graph,
that each of the $n$ populations has its own coalescence rate,
and write $i \sim j$ to mean that $i$ and $j$ are adjacent in the graph.
Then, the resulting log-posterior density is
\begin{align} \label{eq:post}
    \begin{split}
\mathcal{L}(D, G, \gamma) 
    &=
    2 m_G \log(\lambda_G) + n \log(\lambda_\gamma) 
    - \frac{n(n+1)}{2} \log(2 \pi \sigma^2_\epsilon)
	- \sum_{i \sim j} \lambda_G G_{ij}  \\
    &\qquad
    -\sum_{i=1}^n \lambda_{\gamma}\gamma_i
		-\frac{1}{2 \sigma_{\epsilon}^2} \sum_{i \leq j} \left(
            \mathcal{C}_{ij}(G,\gamma) - D_{ij}
        \right)^2 
    \end{split}
\end{align}
Here $1/\lambda_\gamma$ and $1/\lambda_G$ are the prior means of $\gamma_i$ and $G_{ij}$,
respectively.
Recall that both $D$ and $\mathcal{C}(G,\gamma)$ are full matrices of pairwise distances between discrete locations (in the same order), \revpoint{5}{5}
the term $(\mathcal{C}_{ij}(G,\gamma) - D_{ij})^2$
measures the deviation of the observed genetic distances
from those predicted under the coalescent model on the graph. \revpoint{5}{4}

@@

We then test the method's ability to locate barriers to migration
on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
Each nonzero movement rate is determined randomly as before,
and coalescence rates are set to 1 everywhere.
Figure \ref{fig:5x3b_post_coalvcom} shows posterior distributions of the movement rates
for both coalescence time and resistance distance inference
on this graph, with noise standard deviation equal to
1/1000${}^\text{th}$ of the mean coalescence time (top)
and 1/100${}^\text{th}$ of the mean coalescence time (bottom). \revpoint{5}{7}
Although increased noise in the data increased uncertainty in the estimated values,
coalescence time inference correctly inferred not only the location of the barrier
but also the migration rates along all other edges within a reasonable margin of error.
Resistance distance inference also identified the barrier,
but was much less accurate for the other values.
We also explored making the problem harder,
by first (a) dropping data corresponding to populations 2, 5, 7, and 15;
and then (b) allowing a separate coalescence rate for each location,
in both cases using a noise level of 1/100.
The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
and show substantially increased error,
but only in movement rates to the removed locations.
As before, uncertainty was increased when multiple coalescence rates were allowed,
but not as strongly as before, likely because the problem is less close to being underconstrained,
having 105 equations and 59 unknowns.

@@

Most promisingly, the recent method, MAPS \citep[``Migration and Population Surface estimation'',][]{alasadi2018estimating}
is based on EEMS, but instead of resistance distances,
uses pairwise sharing of long haplotype segments,
which have been shown to carry substantial information about recent demography
\citep{ralph2013geography,palamara2013inference,browning2015accurate,ringbauer2017inferring}.
Furthermore, the likelihood model underlying MAPS is based on coalescent theory,
and thus should not be subject to the drawbacks of resistance distance.
\citet{ringbauer2018estimating} also uses shared haplotype lengths
to identify a barrier in a coalescence-based framework.
However, the underlying migration model is symmetric,
and haplotype length-based methods require fairly dense genotyping and a genetic map,
something not available for many species.

@@

All of the methods presented above work with a discretized model of space,
for mathematical and computational reasons. \revpoint{5}{8}
Development of new methods for continuous geography
and assessment of how current methods behave when faced with continuous geography
is an important challenge facing the field.
This has been made substantially easier with the introduction of continuous space 
into the simulator, SLiM \citep{haller2018forward},
that we used to test the behavior of our method with more realistic data.
Another promising approach centers around the Spatial Lambda-Fleming-Viot model
\citep{barton2010continuum}, \revpoint{5}{9}
which provides a mechanistic model in which coalescent simulations are possible.
This has been used for inference \citep{guindon2016demographic},
but more work remains to understand how it behaves as an approximation to real populations.
